# Preface {.unnumbered #sec-preface}

::: {.callout-warning}
## Warning
This is a website for the **work-in-progress** PhD thesis of mine. It is **not** intended to be read by anyone except me *(and maybe few other people)* yet. If you do flick through it anyway, consider yourself warned. It might be messy at some places and will definitely undergo serious rewriting.
:::

::: {.callout-note collapse="true"}
## Note
This work can be read online at <https://petrpajdla.github.io/dataInfrastructures/>. 
The source repository is on GitHub at <https://github.com/petrpajdla/dataInfrastructures/>.
:::

This document is created in an open-source [Quarto](https://quarto.oq()rg/) scientific and technical publishing system. You might be asking why is it published and written like this even if it is not intended for any audiences except myself yet. I have no answer to this. One evening I simply decided to give *Quarto* publishing a try and set this whole thing up in less then an hour or so.

### Notes on writing

This note is written mostly for a future me, in case I need to set up the working environment again on a different machine and to serve as a memo if I forget how to continue.

As of November 2022, this is written on [Archlabs](https://archlabslinux.com/) *GNU/Linux* machine, mostly in [Visual Studio Code](https://code.visualstudio.com/) editor and sometimes in [RStudio](https://posit.co/). Changes are tracked with *Git* and a remote repository is on *GitHub* (see the note above), same as the rendered website. The rendered version of the manuscript is in the branch `gh-pages`. See a guide on how to set this up [here](https://quarto.org/docs/publishing/github-pages.html). The online version is published with this command:

```{.bash filename="Terminal"}
quarto publish gh-pages
```

In my point of view, there are numerous advantages to scientific writing in this manner over traditional *Office*-based approach. A non-exhaustive list of why to do scientific writing this way is below.

- **Plain text**  
  Writing in plain text enhanced with a simple *Markdown* syntax and some *Quarto* elements is great because from one source document, a *.pdf*, *.html*, *.docx* (and probably more) document formats can be rendered using [pandoc](https://pandoc.org/).
- **Version control**  
  Tracking changes using *git* is easily implemented when writing in a plain text. Keeping track of any changes in the manuscript is obviously crucial for any later revisions etc.
- **Simple citation management**  
  Bibliography is organized using [Zotero](https://www.zotero.org/) with [Better BibTeX](https://retorque.re/zotero-better-bibtex/) extension which is used to export (and keep updated) necessary collections in a parent folder of the manuscript as *.bib* files. My *Zotero* library is [here](https://www.zotero.org/knytt). To format the citations, a citation style of the *Journal of Computer Applications in Archaeology* is used (.csl file was obtained [here](https://www.zotero.org/styles/)). 
- **Embedded code**  
  Code blocks (and the associated results) can be easily embedded in the text. My language of choice is *R*. For more information on reproducibility see @marwick2017 and @marwick2018a.


::: {.callout-tip collapse="true"}
## In-text citations
```
@citekey        -> Author (year)
-@citekey       -> (year)
[@citekey]      -> (Author, year)
@citekey [p. X] -> Author (year, p. X)
```
:::

::: {.callout-tip collapse="true"}
## Crossrefs
```
{#sec-label} -> #sec-label
{#fig-label} -> #fig-label
crossref withot numbering: -@sec-label, [Chapter -@sec-label]
```
:::

### Stats

```{r}
#| label: wc-in
#| echo: false
fl <- paste0("chapters/", c("intro.qmd", "theory.qmd", "method.qmd", "data.qmd"))
# count words (w) and characters (c) using shell
system(paste0("wc -wm ", paste0(fl, collapse = " "), " | sed -re 's/ +/,/g' > wc.txt"))
wc <- read.csv("wc.txt", header = FALSE)[, 2:4] |>
  setNames(c("w", "c", "f"))
```

As of `r format(Sys.Date(), "%B %-d, %Y")` there are roughly `r round(wc[nrow(wc), "c"]/1800, 1)` pages of text. 
Length of individual chapters is as follows: 

```{r}
#| label: wc-out
#| echo: false
wc
```


